---
categories:
- ""
- ""
date: "2020-10-20"
description: ""
draft: false
keywords: ""
slug: blog4
title: "Yield Curve inversion"
output: html_document
---



```{r, setup, echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width=6.75, 
  fig.height=6.75,
  fig.align = "center"
)
```


```{r load-libraries, echo=FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(ggthemes)
library(GGally)
library(readxl)
library(here)
library(skimr)
library(janitor)
library(broom)
library(tidyquant)
library(infer)
library(openintro)
library(tidyquant)
```

## Data preparation 

Every so often, we hear warnings from commentators on the "inverted yield curve" and its predictive power with respect to recessions. An explainer what a [inverted yield curve is can be found here](https://www.reuters.com/article/us-usa-economy-yieldcurve-explainer/explainer-what-is-an-inverted-yield-curve-idUSKBN1O50GA). If you'd rather listen to something, here is a great podcast from [NPR on yield curve indicators](https://www.podbean.com/media/share/dir-4zgj9-6aefd11)

In addition, many articles and commentators think that, e.g., [*Yield curve inversion is viewed as a harbinger of recession*](https://www.bloomberg.com/news/articles/2019-08-14/u-k-yield-curve-inverts-for-first-time-since-financial-crisis). One can always doubt whether inversions are truly a harbinger of recessions, and [use the attached parable on yield curve inversions](https://twitter.com/5_min_macro/status/1161627360946511873).

In this case we will look at US data and use the [FRED database](https://fred.stlouisfed.org/) to download historical yield curve rates, and plot the yield curves since 1999 to see when the yield curves flatten. If you want to know more, a very nice article that explains the [yield curve is and its inversion can be found here](https://fredblog.stlouisfed.org/2018/10/the-data-behind-the-fear-of-yield-curve-inversions/). 

First, let us use the `tidyquant` package to download monthly rates for different durations. 

```{r get_rates, warning=FALSE}
# Get a list of FRED codes for US rates and US yield curve; choose monthly frequency
# to see, eg., the 3-month T-bill https://fred.stlouisfed.org/series/TB3MS
tickers <- c('TB3MS', # 3-month Treasury bill (or T-bill)
             'TB6MS', # 6-month
             'GS1',   # 1-year
             'GS2',   # 2-year, etc....
             'GS3',
             'GS5',
             'GS7',
             'GS10',
             'GS20',
             'GS30')  #.... all the way to the 30-year rate

# Turn  FRED codes to human readable variables
myvars <- c('3-Month Treasury Bill',
            '6-Month Treasury Bill',
            '1-Year Treasury Rate',
            '2-Year Treasury Rate',
            '3-Year Treasury Rate',
            '5-Year Treasury Rate',
            '7-Year Treasury Rate',
            '10-Year Treasury Rate',
            '20-Year Treasury Rate',
            '30-Year Treasury Rate')

maturity <- c('3m', '6m', '1y', '2y','3y','5y','7y','10y','20y','30y')

# by default R will sort these maturities alphabetically; but since we want
# to keep them in that exact order, we recast maturity as a factor 
# or categorical variable, with the levels defined as we want
maturity <- factor(maturity, levels = maturity)

# Create a lookup dataset
mylookup <- data.frame(symbol=tickers,var=myvars, maturity=maturity)
# Take a look:
mylookup %>% 
  knitr::kable()

df <- tickers %>% 
  tidyquant::tq_get(get="economic.data", 
                   from="1960-01-01")   # start from January 1960

glimpse(df)
```

Our dataframe `df` has three columns (variables):

- `symbol`: the FRED database ticker symbol
- `date`: already a date object
- `price`: the actual yield on that date

The first thing would be to join this dataframe `df` with the dataframe `mylookup` so we have a more readable version of maturities, durations, etc.


```{r join_data, warning=FALSE}

yield_curve <-left_join(df, mylookup, by="symbol") 
```

## Plotting the yield curve

This may seem long but it should be easy to produce the following three plots

### Yields on US rates by duration since 1960

```{r yield_curve_1, echo=FALSE, out.width="100%", fig.height = 9, fig.width = 12.5}

ggplot(yield_curve, aes(x = date, y = price, color = symbol))+
  geom_line() +
  
  facet_wrap(~var, nrow = 5, ncol = 2, strip.position = "top") +
  
  labs(title = "Yields on U.S. Treasury rates since 1960",
       x = "",
       y = "%",
       caption = "Source: St. Louis Federal Reserve Economic Database (FRED)")+
  
  theme(legend.position = "none",
        panel.background = element_rect(fill = 'white', color = "black", size = rel(0.1)),
        panel.grid = element_line(color = "grey95", size = rel(0.5)),
        strip.background = element_rect(color = "black", size = rel(0.1))
        )
```

### Monthly yields on US rates by duration since 1999 on a year-by-year basis


```{r yield_curve_2, echo=FALSE, out.width="100%", fig.height = 9, fig.width = 12.5}

yield_curve2 <- yield_curve %>% 
  mutate(year_of_date = year(date)) %>% 
  filter(year_of_date > 1998) 


ggplot(yield_curve2, aes(x = maturity, y = price, 
                         color = as.factor(year_of_date), 
                         group = date))+
  geom_line()+
  
  labs(title = "US Yield Curve",
       x = "Maturity",
       y = "Yield(%)",
       caption = "Source: St. Louis Federal Reserve Economic Database (FRED)") +
  
  facet_wrap(~year_of_date, nrow = 6, ncol = 4)+
  
  theme(legend.position = "none",
        panel.background = element_rect(fill = 'white', color = "black", size = rel(0.1)),
        panel.grid = element_line(color = "grey95", size = rel(0.5)),
        strip.background = element_rect(color = "black", size = rel(0.1)),
        text = element_text(size = 8),
        title = element_text(size = 8)) +
  NULL

```



### 3-month and 10-year yields since 1999

```{r yield_curve_3, echo=FALSE, out.width="100%", fig.height = 9, fig.width = 12.5}

yield_curve3 <- yield_curve2 %>% 
  filter(year_of_date > 1998 & (maturity == "10y" | maturity == "3m"))

ggplot(yield_curve3, aes(x = date, y = price, 
                         color = var, 
                         group = var))+
  geom_line()+

  labs(title = "Yields on 3-month and 10-year US Treasury rates since 1999",
       caption = "Source: St. Louis Federal Reserve Economic Database (FRED)",
       x = "",
       y = "%")+
  
  theme(legend.title = element_blank(),
        legend.key = element_blank(),
        panel.background = element_rect(fill = 'white', color = "black", size = rel(0.1)),
        panel.grid = element_line(color = "grey95", size = rel(0.5))) +
  NULL

```


According to [Wikipedia's list of recession in the United States](https://en.wikipedia.org/wiki/List_of_recessions_in_the_United_States), since 1999 there have been two recession in the US: between Mar 2001–Nov 2001 and between Dec 2007–June 2009. Does the yield curve seem to flatten before these recessions? Can a yield curve flattening really mean a recession is coming in the US? Since 1999, when did short-term (3 months) yield more than longer term (10 years) debt?

Besides calculating the spread (10year - 3months), there are a few things we need to do to produce our final plot

1. Setup data for US recessions 
1. Superimpose recessions as the grey areas in our plot
1. Plot the spread between 30 years and 3 months as a blue/red ribbon, based on whether the spread is positive (blue) or negative(red)


```{r setup_US-recessions, warning=FALSE}

# get US recession dates after 1946 from Wikipedia 
# https://en.wikipedia.org/wiki/List_of_recessions_in_the_United_States

recessions <- tibble(
  from = c("1948-11-01", "1953-07-01", "1957-08-01", "1960-04-01", "1969-12-01", "1973-11-01", "1980-01-01","1981-07-01", "1990-07-01", "2001-03-01", "2007-12-01"),  
  to = c("1949-10-01", "1954-05-01", "1958-04-01", "1961-02-01", "1970-11-01", "1975-03-01", "1980-07-01", "1982-11-01", "1991-03-01", "2001-11-01", "2009-06-01") 
  )  %>% 
  mutate(From = ymd(from), 
         To=ymd(to),
         duration_days = To-From) %>% 
  filter(From >= "1960-01-01")

print(recessions)
```

```{r plotting_the_recession, fig.height = 7, fig.width = 10}

spread_plot_long <- yield_curve %>% 
  filter(maturity == "10y" | maturity == "3m") %>% 
  select(date, price, var) 

spread_plot <- spread(spread_plot_long, var, price) %>% 
  mutate(yield_spread = get('10-Year Treasury Rate') - get('3-Month Treasury Bill')) %>% # calculate yield spread   
  mutate(positive_yield = ifelse(yield_spread>0, yield_spread, NA)) %>% # period with positive yield
  mutate(negative_yield = ifelse(yield_spread<=0, yield_spread, NA)) # period with negative yield

ggplot()+
  geom_rect(data = recessions, mapping = aes(xmin = From, xmax = To, ymin = -Inf, ymax = Inf),
            fill = "grey")+ # highlight the recession periods in grey
  geom_line(data = spread_plot, aes(x = date, y = yield_spread))+
  geom_ribbon(data = spread_plot, mapping = aes(x = date, ymin = negative_yield, ymax = 0), fill = "red3", alpha = 0.2)+ # create red ribbon for the negative yield period
  geom_ribbon(data = spread_plot, mapping = aes(x = date, ymax = positive_yield, ymin = 0), fill = "blue3", alpha = 0.2)+ # create blue ribbon for the positive yield period

  labs(title = "Yield Curve Inversion: 10-year minus 3-month U.S. Treasury rate",
       subtitle = "Difference in % points, monthly averages.
Shaded areas correspond to recessions",
       x = "Date",
       y = "Difference (10 year – 3 month) yield in %",
       caption = "Source: St. Louis Federal Reserve Economic Database (FRED)")+
  
  scale_x_date(date_breaks = "2 years", date_labels = "%Y", limits = as.Date(c("1960-01-01", "2020-02-01")))+


  theme(legend.title = element_blank(),
        legend.key = element_blank(),
        panel.background = element_rect(fill = 'white', size = rel(0.1)), # remove panel background
        panel.grid = element_line(color = "grey95", size = rel(0.5)), # adding grid lines
        text = element_text(size = 8),
        title = element_text(size = 12)
        )+ 
        
  NULL
```


Since 1999, the yield curve flattens before both recessions and after both of the yield curve flattening came a recession. There is also a flattening since 2019, yet we could not tell if a recession has come after the flattening.
Since 1999, short-term (3 months) yielded more than longer term (10 years) debt during the following three periods: 
1) from 2000-08 to 2000-12
2) from 2006-08 to 2007-04
3) from 2019-06 to 2019-09

